---
title: "BIOL2010 prac series two"
output: 
  learnr::tutorial:
    progressive: TRUE
    language:
      en:
        button:
          continue: "continue"
          previoustopic: "go back"
          nexttopic: "continue"
runtime: shiny_prerendered
---

```{r setup, include=FALSE}

# attach packages
library(DT)
library(learnr)
library(knitr)
library(readxl) 
library(tidyverse) # always last

set.seed(2010)

# import data -- use point level for workshop
dat <- read_excel("data/BIOL2010-prac-2-2026.xlsx",
                  sheet = "point_level") %>% 
  arrange(ecosystem) %>% 
  add_column(leaf_carbon = c(
    rnorm(n = 53, mean = 20, sd = 7),
    rnorm(n = 48, mean = 60, sd = 7))
  ) %>% 
  mutate(leaf_carbon = round(leaf_carbon, 2)) %>% 
  arrange(transect_id, point)

# and quarter level available for open-ended work
dat_quarters <- read_excel("data/BIOL2010-prac-2-2026.xlsx",
                           sheet = "quarter_level")

# often folks want to twiddle axis text margins
# theme(axis.title.x = element_text(margin = margin(6, 0, 0, 0)))

```

```{r setup-dat-plot, include=FALSE}

dat_plot <- dat %>% 
  group_by(ecosystem) %>% 
  summarise(mean_leaf_carbon = mean(leaf_carbon, na.rm = TRUE),
            se_leaf_carbon = sd(leaf_carbon, na.rm = TRUE)/sqrt(n()))

```

## Visions of data
###
Have a look at your data below! (at the _point-level_ --- what does this mean?)

```{r data-view, echo=FALSE}

datatable(dat,
          rownames = FALSE,
          options = list(dom = "lpt",
                         scrollX = TRUE))

```
<br>
What are two options for visualising this data?

###
Let's first make a ___bar plot___, showing a:

* __categorical predictor__; with a
* __continuous response__.

```{r plot-bar, exercise=TRUE, exercise.setup = "setup-dat-plot"}

ggplot(data = dat_plot,
       aes(x = PRED_VAR)) +
  geom_SOMETHING(aes(y = RESP_VAR),
           fill = "yellow",
           colour = "red",
           width = 0.25) +
  geom_SOMETHING(aes(ymin = SOMETHING,
                     ymax = SOMETHING),
                width = 0.8) +
  scale_x_discrete(name = "something here about ecosystem please",
                   labels = c("mmmmm delicious mud",
                              "pov: you did not have to climb down the slope")) +
  scale_y_continuous(name = "what is basal area??",
                     expand = expansion(mult = c(0, 0.05))) +
  theme_classic()

```

###
Now, let's make a ___scatter plot___, showing a:

* __continuous predictor__; plus a
* __categorical predictor__; with a
* __continuous response__.

```{r plot-scatter, message=FALSE, warning=FALSE, exercise=TRUE}

ggplot(data = dat,
       aes(x = CONT_PRED_VAR, 
           y = RESP_VAR,
           colour = CAT_PRED_VAR)) +
  geom_SOMETHING() +
  geom_smooth(method = "lm",
              se = TRUE, fill = "yellow") +
  scale_colour_discrete(name = "Ecosystem",
                        labels = c("MMMMM angrove",
                                   "SSSSS clerophyll")) +
  # sure would be a shame if someone labelled those there axes...
  theme_classic() 
  
```

You should now have two options for making plots --- including working, modify-able code for each.

## Making a difference
###
For each of our two plots, let's recap how to analyse their data, and interpret the results.

Firstly for our ___bar plot___ research question --- what statistical test is appropriate?

###
We will use an **a**nalysis **o**f **v**ariance.
```{r stat-anova, exercise=TRUE}

bar_plot_vibes <- SOME_TEST(RESP_VAR ~ PRED_VAR,
                      data = dat)

summary(bar_plot_vibes)

```

###
And now for our ___scatter plot___ research question --- what statistical test is appropriate?

###
We will use a **l**inear **m**odel.

```{r stat-lm, exercise=TRUE}

scatter_plot_vibes <- SOME_TEST(RESP_VAR ~ CONT_PRED_VAR * CAT_PRED_VAR,
                         data = dat)

summary(scatter_plot_vibes)

```

## Time to tidy up
###
Wrangling the data will always be your first step in R. We skipped to the second and third steps of visualising and analysing to start the class. Time to catch up on forgotten fun!

Let's practice some example data tidying together. 

###
__Making means and SEs for bar plots:__ `group_by()` and `summarise()` work together for this. 

```{r tidy-summarise, message=FALSE, exercise=TRUE}

dat_plot <- dat %>% 
  FUNCTION_1(PRED_VAR) %>% 
  FUNCTION_2(SOMETHING_ABOUT_RESP_VAR)

# you may need to use 'na.rm = TRUE' to avoid
# issues with missing data

dat_plot

```

###
__Selecting data from one ecosystem:__ `filter()` is a good option, and maybe you'd also like to `select()` just your variables of interest. 

```{r tidy-filter, message=FALSE, exercise=TRUE}

mang_dat <- dat %>% 
  FUNCTION_1(SOMETHING_ABOUT_ECOSYSTEM) %>% 
  FUNCTION_2(SOME_VARS_YOU_WANT)

mang_dat

```

###
__Creating a new variable:__ `mutate()` is very flexible, and fits nicely into a pipeline. 

```{r tidy-mutate, message=FALSE, exercise=TRUE}

mang_dat <- dat %>% 
  filter(ecosystem == "mang") %>% 
  FUNCTION_1(SOMETHING_FORMULAIC)

# let's say you found an allometric equation that 
# claimed that tree weight (in kg) was equal to:
# basal area cubed, divided by 10

mang_dat

```

## Werk werk werk werk werk

Use this working space to design a figure and analysis to complete your practical journal for this week. Use your code from earlier in the tutorial!

Recall that two datasets are available to you by default:

* `dat` --- the _point-level_ dataset.
* `dat_quarters` --- the _quarter-level_ dataset. 

Need a refresher on the difference? Call them by name below and have a look!

```{r working-space, message=FALSE, exercise=TRUE}

# 1. tidy data
# [your code here!]

# 2. visualise data
# [your code here!]

# 3. analyse data
# [your code here!]

```
